---
title: "Clustering Results"
author: "Edward Gregr"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    keep_tex: true       # Keep intermediate LaTeX file for troubleshooting
    latex_engine: pdflatex # Specify the LaTeX engine (pdflatex, xelatex, or lualatex). This renders the pdf
    number_sections: true # Optional, if you want numbered sections
    toc: true             # Table of contents (optional)
    fig_caption: true     # Enable figure captions
fontsize: 11pt            # Set font size (optional)
geometry: margin=1in      # Set margins (optional)
header-includes:
  - \usepackage{booktabs} # for tables
  - \usepackage{pdflscape}
  - \usepackage{tocloft}
  - \usepackage{float} # for controlling placement of tables and figures
  - \usepackage{placeins} # for a cmd to keep text in place.
  - \setlength{\intextsep}{5pt} % Vertical space above & below [h] floats
  - \setlength{\textfloatsep}{5pt} % Vertical space below (above) [t] ([b]) floats
  - \setlength{\abovecaptionskip}{5pt}
  - \setlength{\belowcaptionskip}{5pt}
---

# Summary
This report summarises the clustering of FVCOM point data to support the development of the 
Broughton LSSM. Major challenge is that the resolution of the FVCOM model does not 
support the resolution we are interested in. Lots of geography missing. So, need to 
develop a clear set of objectives. 

# Objectives 
Primary objective is to classify the FVCOM data in a way that can be informative to the LSSM. 
This includes winter/summer classification, and surface and deep classes. So 4 individual classifications.

Secondary objective is to produce a report in an RMD that compares the clusterings. This
can be a bit complicated particularly for the data prep portion, as different predictors - by definition 
since the FVCOM output is divided into JAN/JUL and Surf/Bott. 

# Approach
You have also just (Sep 3) copied some code into the functions script to match up clusters. 
While this is a post-hoc thing, it will need testing.


# Results 
When examining cross-correlations, currents tended to be correlated with each other, as 
well as with chemistry. From a total of 36 predictors (across 2 depths, 2 seasons, and 3 aggregations),
11 were found to be correlated at < 0.60. 

```{r Histograms, results='asis', echo=FALSE }
hist1
hist2
hist3
```

## Seasonal comparison
There are 4 predictors left for winter (DEC). Scree plot suggests up to 9 clusters. No clear 'break' in the curve. 
There are 7 predictors left for summer (JUL). Scree plot is similar to winter, in that up to 9 clusters seem warranted.



```{r, ScreePlotDescripTable, echo=FALSE, escape=FALSE, fig.pos='H'}
#Only needed once per number of variables. 
scree_plot

```

```{r Plots, results='asis', echo=FALSE }

heat_map

v_plots

hist( cluster_result$cluster )

clust_map

```


<!-- THE END -->
