---
title: "Clustering Results"
author: "Edward Gregr"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    keep_tex: true       # Keep intermediate LaTeX file for troubleshooting
    latex_engine: pdflatex # Specify the LaTeX engine (pdflatex, xelatex, or lualatex). This renders the pdf
    number_sections: true # Optional, if you want numbered sections
    toc: true             # Table of contents (optional)
    fig_caption: true     # Enable figure captions
fontsize: 11pt            # Set font size (optional)
geometry: margin=1in      # Set margins (optional)
header-includes:
  - \usepackage{booktabs} # for tables
  - \usepackage{pdflscape}
  - \usepackage{tocloft}
  - \usepackage{float} # for controlling placement of tables and figures
  - \usepackage{placeins} # for a cmd to keep text in place.
  - \setlength{\intextsep}{5pt} % Vertical space above & below [h] floats
  - \setlength{\textfloatsep}{5pt} % Vertical space below (above) [t] ([b]) floats
  - \setlength{\abovecaptionskip}{5pt}
  - \setlength{\belowcaptionskip}{5pt}
---

# Summary
This report summarises the clustering of FVCOM point data to support the development of the 
Broughton LSSM. Major challenge is that the resolution of the FVCOM model does not 
support the resolution we are interested in. Lots of geography missing. So, need to 
develop a clear set of objectives. 

# Objectives
Develop a seasonal (winter/summer) classification relevant to bull kelp using the native resolution of the FVCOM data.
The expectation is that this could lead to a better understanding of how and why bull kelp distributions have been limited recently in the area. 

Secondary objective is to produce a report in an RMD that compares the seasonal clusterings. 

# Methods
## Data preparation
We obtained FVCOM data for the region, summarized as mean, min, and max for the representative months of December and July. Ocean variables included currents, temperature, and salinity. Statistics were calculated at the surface and the bottom of each model node. Current values were interpolated from the model edges to the nodes. 

## Variable selection
We first identified predictors likely to be relevant for each season based on kelp biology. We then examined the seasonal sets of predictors and removed highly correlated variables, ensuring that there were no cross-correlations greater than 0.60.

For summer, we prioritized monthly surface temperatures, preferentially retaining the mean followed by the maximum as waters are generally still cool (i.e., unlikely to have reached critical temps REF). For salinity, we prioritized bottom salinity, also preferring the mean followed by the maximum to emphasize areas of well mixed, nutrient-rich ocean waters. We retained current predictors only if they were correlated at less than 0.6 with temperature or salinity, again preferentially retaining the mean followed by the maximum to emphasize areas of mixing.

For winter, given the importance of gametophyte settlement and transport, we prioritized mean bottom currents, temperature, and salinity, preferring means followed by minimum values. 

# Results 
## Correlation analysis






When examining cross-correlations, currents tended to be correlated with each other, as 
well as with chemistry. From a total of 36 predictors (across 2 depths, 2 seasons, and 3 aggregations),
11 were found to be correlated at < 0.60. 

```{r Histograms, results='asis', echo=FALSE }
hist1
hist2
hist3
```

## Seasonal comparison
There are 4 predictors left for winter (DEC). Scree plot suggests up to 9 clusters. No clear 'break' in the curve. 
There are 7 predictors left for summer (JUL). Scree plot is similar to winter, in that up to 9 clusters seem warranted.



```{r, ScreePlotDescripTable, echo=FALSE, escape=FALSE, fig.pos='H'}
#Only needed once per number of variables. 
scree_plot

```

```{r Plots, results='asis', echo=FALSE }

heat_map

v_plots

hist( cluster_result$cluster )

clust_map

```


<!-- THE END -->
