---
title: "Coastal clusters for the Local Seaweed Services Model"
author: "Edward Gregr"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    keep_tex: true       # Keep intermediate LaTeX file for troubleshooting
    latex_engine: pdflatex # Specify the LaTeX engine (pdflatex, xelatex, or lualatex). This renders the pdf
    number_sections: true # Optional, if you want numbered sections
    toc: true             # Table of contents (optional)
    fig_caption: true     # Enable figure captions
fontsize: 11pt            # Set font size (optional)
geometry: margin=1in      # Set margins (optional)
header-includes:
  - \usepackage{booktabs} # for tables
  - \usepackage{pdflscape}
  - \usepackage{tocloft}
  - \usepackage{float} # for controlling placement of tables and figures
  - \usepackage{placeins} # for a cmd to keep text in place.
  - \setlength{\intextsep}{5pt} % Vertical space above & below [h] floats
  - \setlength{\textfloatsep}{5pt} % Vertical space below (above) [t] ([b]) floats
  - \setlength{\abovecaptionskip}{5pt}
  - \setlength{\belowcaptionskip}{5pt}
---

\newpage

\listoftables
\listoffigures

\newpage

# Introduction
This work follows the approach described in Gregr (2024a) where k-means is applied to DFO MSEA data to create a spatial structure to support marine use planning initiatives. In the absence of an empirical representation for data-poor species, clusters describing areas of similar environmental conditions to be used as a framework to collate existing information on the distribution of kelp and other nearshore benthic species. 

This application uses oceanographic predictors output from an ocean circulation model (Bianucci et al. 20XX). These predictors describe average ocean conditions for the month of July. 

The code assumes the TIFs are in the same projection, but imposes a common extent and resolutionto ensure all inputs have identical spatial configurations. 

See Gregr (2024a) for details on the methods and the k-means classification.

# Methods
After standardizing the spatial representation of the source data, the data are normalized and scaled, and clipped to a reduced spatial extent if desired. Correlations are examined and Where predictors are cross-correlated > 0.6, one of them is removed.

Clusters are then generated from samples, as the data can be quite large. This can lead to the emergence of different clusters unless, as here, a seed is specified. However, the clusters still end up different each time as the algorithm itself chooses a new starting point at random for each cluster whenever its run. Final clusters can be created using the full data set. 

The working clusters are  then examined in a variety of ways including:
1.	A heat map of the within cluster sum of squares. 
2.	Silhouette plots of the clusters 
Silhouette plots provide information on how observations fit their clusters. Generally, the higher the average width of each clusters the better. 
3.  PCA plots of the clusters
4.	violin plots showing the predictor contributions to each cluster


# Data sources
The output from the Bianucci model has been pre-processed by Barbosa (Table 1). 

```{r, DescripTable, echo=FALSE, escape=FALSE}
library(knitr)

descrip_table <- data.frame(
  Process = c("Light", "Energy", "Exposure", "Surface Salinity, Bottom Salinity", "Surface temperature, Bottom temperature"),
  Predictor = c("Northness", "tidal_cur, julSSpd_ave, julBSpd_ave", "Wind", "julSS_min, julSS_ave, julBS_ave", "julST_ave, julST_max, julBT_ave, julBT_max"),
  Description = c(
    "Kelps are light-restricted, so northness provides a measure of available sunlight.",
    "Energy provides an indication of water flow and nutrient mixing.", 
    "Exposure can influence substrate and also provide and indicator of mixing",
    "Kelps do better in higher salinity waters, as salinity tends to be correlated with both cooler temperatures and increased nutrients.",
    "Kelps do better in cooler waters"
  )
)

kable(descrip_table, format = "latex", booktabs = TRUE, 
      caption = "The predictors available from DFO’s MSEA group for different potential drivers of kelp habitat suitability, and the rationale for their inclusion.")  %>%
  kable_styling() %>%
  column_spec(1, width = "1.5in", latex_column_spec = "p{1.5in}") %>% 
  column_spec(2, width = "1in", latex_column_spec = "p{1in}") %>% 
  column_spec(3, width = "3.6in", latex_column_spec = "p{3.6in}")     

```

As part of the pre-processing, Barbosa et al. removed unsuitable depths from the source data to focus the clustering on the photic zone. 

*Add substrate to remove soft and sand substrate*

# Results 

## Data description

The loaded Raster stack is `r dim(selected_stack)[1]` by `r dim(selected_stack)[2]`, giving a total of `r scales::comma( dim(selected_stack)[1] * dim(selected_stack)[2])` pixels in the study domain. Much of this area is land, or deeper waters excluded by the bathymetry.

*Include: proportion nearshore, proportion of that with good substrate, and final sample size for clustering.* 

## Predictor assessment

### Cross-correlations
When cross-correlations exceeded 0.6 among the predictors (Table 2, Table 3), one of them was removed. Cross-correlations included: *julSSpd_ave* with *julBSpd_ave*, *julSS_ave* and *julSS_min*, and *julST_ave* and *julST_max*. Current speed at the surface was deemed more important than at the  bottom; minimum salinity more important than average salinity, and maximum SST more important than average. The predictors *julST_ave*, *julSS_ave*, *julBSpd_ave* were therefore dropped. Additionally, after initial explorations of clusters, *northness* was seen to constantly contribute equally across clusters and was thus also dropped. 

This left **8** potential predictors.

```{r Correlations, results='asis', echo=FALSE, table.pos='t' }
# Caption is part of kable()
# Using global: cor_table 

y_low <- lower.tri(cor_table, diag = TRUE)
cor_table[ y_low ] <- NA
knitr::kable( cor_table, digits = 2, format = "latex", booktabs = TRUE, 
              caption = "Correlation matrix for assessing predictor cross-correlations") %>%
  kable_styling( latex_options = c("scale_down") ) %>% 
  row_spec(0, angle = 90) 
#%>% landscape()

x <- 0.6
high_rows <- apply(cor_table, 1, function(row) any(row > x, na.rm = TRUE))
z <- cor_table[ high_rows, ]
knitr::kable( z, digits = 2, format = "latex", booktabs = TRUE, 
              caption = "Predictor variables that exceed 0.6 threshold") %>%
  kable_styling( latex_options = c("scale_down") ) %>% 
  row_spec(0, angle = 90) 
#%>% landscape()

```
\newpage
## Distributions and outliers
Skewness scores showed strong nonnormality for 4 predictors. Transforms were applied to *julBS_ave*, *julSS_min*, *julSSpd_ave*, and *tidal_cur*. Simple transforms were effective at reducing skewness to less than |1|, thus ceilings were not required (**Table 4**). 

```{r SkewTable, results='asis', echo=FALSE, table.pos='t'}

skdat <- data.frame(Predictor = character(), Pre_Skew = numeric(), Post_Skew = numeric(), stringsAsFactors = FALSE)

# Loop through the columns of the dataset
for (i in 1:ncol(t_stack_data)) {
  # Calculate skewness for each column in both datasets
  pre  <- skewness(stack_data[, i], na.rm = TRUE) 
  post <- skewness(t_stack_data[, i], na.rm = TRUE) 
  
  # Combine the column name and skewness values into a new data frame row
  new_row <- data.frame(
    Predictor = colnames(t_stack_data)[i],  # Get column name
    Pre_Skew  = round(pre, 3),                     # Skewness before transformation
    Post_Skew = round(post, 3)                     # Skewness after transformation
  )
  # Append the new row to skdat
  skdat <- rbind(skdat, new_row)
}

# Now add the transform values 
# Create the data frame
# df <- data.frame(
#   Predictor = c("REI", "freshwater_index", "standard_dev_slope", "temp_range"),
#   Ceiling = c(0.3, 0.025, 10, NA),
#   Power_transform = c("1/2", "1/3", "1/2", "1/2")
# )
# Merge the data frames


# Print the resulting data frame
knitr::kable( skdat, format = "latex", booktabs = TRUE, 
              caption = "Transforms applied to skewed predictors and the resulting change in skewness.") %>%
kable_styling(latex_options = "hold_position")

```

Pre (Figure 1) and post (Figure 2) histograms showing the effects of the transformation and scaling of the data. 
\newpage
```{r preHists, echo=FALSE, fig.pos='h', fig.cap="Histograms of the selected, unmodified predictors.", fig.align='center', fig.width=8, fig.height=3.75}

# fig.pos='H', 
#```{r PreAndPostHists, echo=FALSE, include=FALSE, fig.pos='t', fig.cap= c("Histograms of the #selected, unmodified #Predictors.", "Histograms of selected, transformed, and scaled predictor #data."), out.extra='keepaspectratio', Fig.align='center'}

# Plot the first group of 8 histograms
par(mfrow = c(2, 4), mar = c(2, 2, 2, 2))

for (i in 1:dim(stack_data_clean)[2]) {
  hist(stack_data_clean[, i], nclass=50, main = colnames(stack_data_clean)[i], xlab="")
}

```

<!-- \smallskip 
     \medskip -->
\bigskip

```{r transHists, echo=FALSE, fig.pos='b', fig.cap="Histograms of the selected, transformed, and scaled predictor data.", fig.width=8, fig.height=3.75}

# Plot the second group of 8 histograms
par(mfrow = c(2, 4), mar = c(2, 2, 2, 2))

for (i in 1:dim(t_stack_data)[2]) {
  hist(t_stack_data[, i], nclass=50, main = colnames(t_stack_data)[i], xlab="")
}

```
\newpage
\FloatBarrier

# Cluster exploration

The number of clusters is informed by a scree plot (Figure 3). This plot compares the total within-cluster sum of squares (TWSS) for an increasing number of clusters. Scree plots show how the TWSS is reduced with each additional cluster. The optimal number of clusters is found near the elbow in the data (i.e., the point beyond which the reduction of TWSS becomes small with each additional cluster.

```{r ScreePlot,  warning=FALSE, message=FALSE, echo=FALSE, fig.cap="Scree plot showing the total within sum-of-squares across a range of cluster numbers."}
plotme
```

Repeated scree plots generated using subsamples (n=50,000) of complete cases provide more information than a scree plot of all the data. Based on a manual assessment of repeated plots, the number of clusters before the breakpoint varied between 6 and 10. The scree plot of all the data breaks at 8 (Figure 3).

Heat maps of the cluster standard deviations (Figure 4) show how the various predictor values are distributed among the clusters. 
```{r HeatMap, warning=FALSE, message=FALSE, echo=FALSE, fig.cap="Heat map showing within cluster standard deviation of the predictors.", fig.show='asis', fig.align='center',fig.width=6, fig.height=4}
z_heat
```

Silhouette plots show the width of the clusters and the total number of values (i.e., pixels) within each (Figure 5). Higher average silhouette width indicates a better fit with the number of clusters. 
```{r SilhouettePlot, warning=FALSE, message=FALSE, echo=FALSE, fig.cap="Silhouette plot showing pixel membership in each cluster and silhouette widths.", figure.placement='H', fig.align='center',fig.width=6, fig.height=4}
par(mar = c(2, 2, 2, 2)) 
plot(sk, col = 1:nclust, border=NA, main = "" )
```

\newpage
## Part 2 - Clusters and predictor loadings
While exercising the cluster analysis, northness was found to contribute equally to the clusters regardless of number of clusters. It was dropped to simplify the analysis. 

The PCA plots (Figure 6, Figure 7) show how the clusters are distributed across the first two (Figure 7) and second two (Figure 8) dimensions.

The loadings are quantified in Table 5, and shown graphically in Figure 8. 

\FloatBarrier
```{r PCATable, warning=FALSE, message=FALSE, echo=FALSE}

loads <- round( pca_results$loadings$rotation, 3 )
knitr::kable( loads, format = "latex", booktabs = TRUE, 
              caption = "Correlation matrix for assessing predictor cross-correlations") %>%
  kable_styling( latex_options = c("scale_down") )

```

\newpage
```{r PCAPlot1, warning=FALSE, message=FALSE, echo=FALSE, fig.cap="PCA Plots showing the clusters across the first and second dimensions.", fig.pos='t', fig.align='center',fig.width=6.5, fig.height=3.7}
par(mar = c(0, 0, 0, 0)) 
plot( pca_results$plot1 )
```

```{r PCAPlot2, warning=FALSE, message=FALSE, echo=FALSE, fig.cap="PCA Plots showing the clusters across the third and fourth dimensions.", fig.pos='b', fig.align='center', fig.width=6.5, fig.height=3.7}
par(mar = c(0, 0, 0, 0)) 
plot( pca_results$plot2 )
```

```{r ViolinPlot, warning=FALSE, message=FALSE, echo=FALSE, fig.cap="Violin plots showing distribiton of predictors in each of the k-means clusters."}
vplots
```

<!-- THE END -->